A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-08-24 main.tex
4.4 edits - in progress

In section \ref{sec:rak-ggmix} we briefly mentioned the existence of two LMM-Lasso variants: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. As we will see, the two generally perform quite similarly. Thus, the preceding results, although based on the LMM-Lasso-Rakitsch implementation, are representative of both LMM-Lasso methods. The primary purpose of this review is to compare LMM-Lasso to other penalized regression methods with respect to correcting for population structure and environmental confounding. However, given the generally strong performance of LMM-Lasso, we also wish to discuss the key differences between these two implementations and provide guidance for potential users.

LMM-Lasso-Rakitsch estimates the variance components once, under the assumption of a null model. LMM-Lasso-ggmix seeks to improve upon this by re-estimating the variance components and $\bbeta$ iteratively. Although this seems a reasonable approach, we have been unable to find a scenario where LMM-Lasso-ggmix outperforms the LMM-Lasso-Rakitsch in estimation accuracy of $\bbeta$. Figure \ref{fig:eta_beta_mse} shows the MSE for estimation of both $\bbeta$ and $\eta$ in the presence of varying $\eta$ values and sample sizes, $n$, for both LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. The bottom panel of this figure shows a modest improvement in the estimation of $\eta$ for the LMM-Lasso-ggmix method, particularly in the presence of lower $\eta$ values. However, these apparent improvements in estimating $\eta$ do not improve the $\bbeta$ estimation accuracy of LMM-Lasso-ggmix. This is evidenced by the top panel of Figure \ref{fig:eta_beta_mse}, which shows nearly identical performances of both methods in terms of $\bbeta$ MSE.

\anna{add some info/figure on convergence and number of iterations for lmm-lasso vs ggmix, delete the following time figures and references to them}

@@

In addition, we have found that LMM-Lasso-ggmix performs poorly in a low-dimensional setting due to the ill-conditioning of the RRM. Given its intended use for the analysis of high-dimensional data, this fact could be of little consequence to practitioners. However, although its estimation accuracy in high dimensions is on par with that of LMM-Lasso-Rakitsch, as shown in Figure \ref{fig:eta_beta_mse}, this is also a setting in which the computational time required may be prohibitive for practical applications. \anna{time interpretation depends on reason ggmix is slow...} Figure \ref{fig:eta_beta_mse} shows the ratio of computing time required for LMM-Lasso-ggmix compared to that of LMM-Lasso-Rakitsch, for various sample sizes. The horizontal dashed line at 1 indicates both methods require the same computing time. Despite its comparable performance in accurately estimating $\bbetaHat$, LMM-Lasso-ggmix takes requires an average of \anna{73.4} times the computing time of LMM-Lasso-Rakitsch. Thus, LMM-Lasso-Rakitsch seems to be a more attractive and scalable method than LMM-Lasso-ggmix.

>>>>

In section \ref{sec:rak-ggmix} we briefly mentioned the existence of two LMM-Lasso variants: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. Both methods perform quite similarly in terms of $\bbeta$ MSE and variable selection. Becasue of this, the preceding results, though generated using the LMM-Lasso-Rakitsch implementation, are representative of both LMM-Lasso methods in the $n = 200, p = 1000$ setting. The primary purpose of this review is to compare LMM-Lasso to other penalized regression methods with respect to correcting for population structure and environmental confounding. However, given the generally strong performance of LMM-Lasso, we also wish to discuss the key differences between these two implementations and provide guidance for potential users.

LMM-Lasso-Rakitsch uses a two-step approach. First, the variance components are estimated under the assumption of a null model. Then, the estimated variance components are used to transform the data, and the $\bbeta$ are estimated using coordinate descent. LMM-Lasso-ggmix seeks to improve upon this procedure by re-estimating the variance components and $\bbeta$ iteratively. Although this seems a reasonable approach, we have been unable to find a scenario where LMM-Lasso-ggmix outperforms the LMM-Lasso-Rakitsch in estimation accuracy of $\bbeta$. In order to systematically compare both methods, we carried out simulations where $p = 1000$, $n = \{200, 400, 600, 800, 1000, 1200, 1400\}$, and $\eta = \{0.2, 0.5, 0.8\}$. Figure \ref{fig:eta_beta_mse} shows the MSE for estimation of the parameters $\bbeta$ and $\eta$, in addition to the corresponding computing time required.

LMM-Lasso-Rakitsh and LMM-Lasso-ggmix achieve nearly identical MSE for $\bbeta$ in many scenarios, particularly where $n << p$, as shown in the top panel of Figure \ref{fig:eta_beta_mse}. When $n \ge p$ the estimation performance of LMM-ggmix becomes erratic, and in some cases worse than that of LMM-Lasso-Rakitsch. This is most dramatic in the presence of substantial environmental environmental confounding, $\xi = 0.8$, and $n = \{1200, 1400\}$. LMM-Lasso-ggmix achieves a moderate improvement in the estimation of $\eta$, particularly in the presence of lower $\eta$ values and where $n << p$, as shown in the middle panel of Figure \ref{fig:eta_beta_mse}. Interestingly, these improved $\eta$ estimates do not correspond to smaller $\bbeta$ MSE for LMM-Lasso-ggmix. When $n \ge 1000$, LMM-Lasso-ggmix's estimation of $\eta$ in suffers. When $n \ge 1000$ and in the presence of substantial environmental confounding, $\xi = 0.8$,  LMM-Lasso-ggmix's is hindered to the point that LMM-Lasso-Rakitsch estimates $\eta$ more accurately.

Overall these results show that LMM-Lasso Rakitsch outperforms LMM-Lasso-ggmix in estimating $\bbeta$ and $\eta$ where $n \ge p$, likely due to the ill-conditioning of the RRM in these settings \anna{cite French paper (?)}. 
% Given the intended use of these methods for the analysis of genetic data, where the number of genetic variants often outnumber the number of subjects, these results could be of little consequence to practitioners. 
When $n < p$, LMM-Lasso-ggmix estimates $\bbeta$ as well as LMM-Lasso-Rakitsch, and $\eta$ better than LMM-Lasso-Rakitsch. However, in this setting the computational time LMM-Lasso-ggmix reqiures may be prohibitive for practical applications. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required for each method in the various simulation settings.

\anna{ below in progress}

Unfortunately, the settings where ggmix estimates things well coincide with excessive times and when it works well in terms of time, it estimates poorly

Despite its comparable performance in accurately estimating $\bbetaHat$, LMM-Lasso-ggmix takes requires an average of \anna{73.4} times the computing time of LMM-Lasso-Rakitsch. Thus, LMM-Lasso-Rakitsch seems to be a more attractive and scalable method than LMM-Lasso-ggmix.

@@

\anna{add some info/figure on convergence and number of iterations for lmm-lasso vs ggmix}
